Manuscript

### 2018-04-27 manuscript/manuscript.Rmd
updated methods

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone triangulation.

However, not all GPS measurements are made equal. It is useful to classify GPS logs into different categories depending on their origin because different types of logs pose different methodological problems. We argue that social science researchers should explore using automatically collected smartphone GPS logs.  In most studies participants receive a specialised GPS devices to track their movement. We call resulting logs *specialised logs*.

@barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). The problem with this is that researchers have to develop a custom-made app, distribute it among research participants and enforce participation. Participants tend not to like these tracking apps because they drain the battery of the device.

Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints ( these constraints will be detailed subsequently). These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with missing data and measurement noise.

| Feature        | Specialised Logs           | Custom logs  | Secondary logs |
| ------------- |:-------------:| -----:| -----:|
| Sampling Frequency | Regular | Regular |  Irregular |
| Sampling Frequency (Hz) | 0.1>      | Between 0.1 and 0.01| <0.01 |
| Typical Duration | 7 days    |    7-14 days | Years |
| Typical Accuracy Variance | Low    |    Medium | High |

@@

include_graphics("img/missingBoaz5minExample.png")
```

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are autocorrelated in both time and space. This means that common methods, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers. Android phones collect location information through a variety of methods.

*removed somethign here*

 In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

Google location history provides a measure of accuracy that is given in meters such that it represents the radius of a 67% confidence circle. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Inaccurate measures are often followed by more accurate measures.  Most notably, low accuracy often (but not always) is  associated with movement (\@ref(fig:accuracyPlot2)). Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. ",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. All journeys are marked with a red line. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. All other journeys are not shown in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

@@

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. This is known as data drift (REFERENCE). If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure 1. The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

@@

This section aims to explore related work with regard to missing data in GPS logs. We will argue three points. First, there is a lack of established practices to follow with respect to missing data in human mobility logs. Second, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs. Third, we discuss in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; **WU**].

How have researchers dealt with missing data in GPS-based human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. Ostensibly this is because they are unaware of possible solutions. Often researchers simply downsample and remove missing observations or use some sort of rule-based common sense imputations (e.g. @palmius_detecting_2017). The only principled approach that we know of that aims to solve the issue of missing data in GPS logs as they relate to human mobility is of @barnett_inferring_2016. Both of these methods will be detailed subsequently, after introducing exploring other spatiotemporal methods.

@@

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance, assumptions which are violated in the case of human movement patterns. For this reason we do not consider SSMs to be useful for imputing missing data in this case.

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal autocorrelations. For instance, the CUTOFF method estimates missing values using the nearest observed neighbours [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; **WU**]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm.

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by:

@@

On the other hand, **Wu 2017** use what they call a Spatial Temporal Semantic Neural Network to predict future human movement. While the authors are concerned with prediction and not imputation, they devised an algorithm called the Spatial Temporal Semantic algorithm converts raw measurements to machine learning friendly discrete bins. This minimises the measurement error and paves the way for applying machine learning methods to human mobility problems.

@@

Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidian plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ... < t_{n+1}$. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

@@

### MYMETHOD algorithm

Our algorithm is designed to leverage the periodic nature of human movement along with the long span of the data to deal with measurement sparcity and inaccuracy.

@@

### MYMETHOD algorithm

@@

1. Extract from measurements $D$ all pause locations and path clusters. Following **Wu 2017 (NOTE: ADD REFERENCE)** we use a modified spatial-temporal-semantic (STS) feature extraction algorithm to transform pause and path locations into machine learning friendly discrete location sequences.
2. Assign each measurement $D$ to a unique discrete location. 
3. Use neural networks to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.

In this paper we have chosen relatively simple functions as component functions of MYMETHOD. It is clear each of them can be improved. We aim to illustrate not the efficacy of the components themselves but rather how they can work together to impute missing data.

#### Extract pause locations

There are multiple ways of extracting pause locations in the literature (**examples & citations needed: see Wu 2017**). As far as we know there are only two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these processes pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. In particular, @barnett_inferring_2016 use a version of @rhee_human_2007's rectangular method to classify pauses and determine flight sub-sections. MYMETHOD uses a stricter version of @barnett_inferring_2016's approach to extract pauses.

@@

1. The next measurement $D_{t+1}$ is within $t_{\text{Pause lim}}$ amount of seconds (so it is not missing)
2. The next measurement $D_{t+1}$ is within $d_{\text{Pause lim}}$ meters.
3. The duration of the pause is more than $\delta_{\text{Pause lim}}$ seconds.
4. Let the points of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$. These points are only a pause if the distance between the mean coordinates of $D_{t,t+1,...,t+n}$ and the furthest away points of $D_{t,t+1,...,t+n}$ is within 2 times the mean accuracy $a$ of $D_{t,t+1,...,t+n}$.

@@

#### Extract paths

To extract path clusters we apply a modified version of the spatial-temporal-semantic (STS) feature extraction presented in **Wu 2017 (NOTE: ADD REFERENCE)**.

The key modification is that whilst **Wu 2017 (NOTE: ADD REFERENCE)** use map matching as a ground truth to assign noisy measurements into discrete bins along paths, we use the total mobility history of the individual to do so. This leverages the periodic nature of human movement as well as the long span of secondary logs. Therefore we consider each measurement at $D_x$ as a sample of $Path_{ab}$. In practice, for each observed path, MYMETHOD uses the following algorithm to estimate $Path_{ab}$:

1. Take all measurements which are not pauses, filter them based on an accuracy threshold $a_{\text{Path Lim}}$.
2. Create a distance matrix for each point and hierachically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.

At this point all empirically observed path clusters and pause clusters are extracted. However, there may be some overlap between pause clusters and path clusters. Thus, the clusters are clustered again, such that the pause clusters retain priority. This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed. The reasoning for this is twofold: first, that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster. Secondly, due to data drift it can often appear that there is movement when there is none.The end result is a discretised map which contains pause and flight clusters based on the history of the individual.

Subsequently, all measurements $D$ were are then assigned to a discrete cluster. We used a simple assigning function, whereby the measurements where assigned to the cluster nearest to the measurement. We shall discuss the merits of choosing a more sophisticated assigning function subsequently.

#### Neural network

An artificial neural network was trained to predict the missing values. The input variables to the neural net are:

1. The previous and subsequent observed cluster.
2. The distance in time to the next & previous cluster.
3. The time of the day encoded as a cyclical two-dimensional feature **NOTE: add reference**
4. The day of the month encoded as a cyclical two-dimensional feature **NOTE: add reference**
5. The month of the year as a binary class matrix.

@@

For a missing timepoint at $D_t$, the output of the model is a set of probability estimates associated with every location cluster. That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per cluster) associated with where the individual is.

>>>>

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone tower triangulation. Smartphone location logs contain measurements from all of these sources, usually in the form of timestamped lattitude, longitude and accuracy values. The accuracy $a$ of any given measurement is given in meters such that it represents the radius of a 67% confidence circle [@location_history_timeline_2017]. In other words, the true location of a device should be within the radius $a$ of the measurement 67% of the time.

Researchers often develop custom-made tracking applications which participants are instructed to download on their phone. Alternatively participants are given a phone to use for a given period of time with the custom-made tracking app pre-installed. We call location logs resulting from these custom-made apps *custom logs*. The advantage of custom logs is that the researchers can adjust tracking parameters, such as the the frequency of measurements and the sensor with which they are made. The disadvantage with this approach is that researchers have to develop or adapt a custom-made tracking application (which is not easy given hundreds of different types of smartphone models), distribute it among research participants and enforce participation. Participants may dislike tracking apps because they view them as more intrusive and these apps regularly drain the battery of the device [@harari_using_2016]. Moreover, researchers have distribute this application among research participants and convince them not to turn it off.

We focus on another solution, which is to take advantage of existing smartphone location logs  (*secondary logs*) . The advantages are clear: repositories such as Google's Location History contains information on millions of users spanning years [@location_history_timeline_2017], participants can share the data by the click of a button and there can be no behavioural changes due to participation in the study as the participant share past data. The disadvantage is that researchers have no control over the tracking parameters, often resulting in logs with sparse and inaccurate measurements. Hence, two important challenges are dealing with missing data and measurement noise.

@@

include_graphics("img/missingBoaz5minExample.png")
#change y axis lables to natural numbers, x-axis lables to shorter periods and colour to shapes.

@@

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are autocorrelated in both time and space. This means that best practices with other types of data, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work, even though she has never been there. Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS location trackers because smartphones often use less accurate sensors. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time. Figure \@ref(fig:accuracyPlot) shows how accuracy can vary as a function of user behaviour, time and location. Most notably, low accuracy is often (but not always) associated with movement (\@ref(fig:accuracyPlot2)).

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. @palmius_detecting_2017 note that in smartphone custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. Inaccurate measures are often followed by more accurate measures. There are several recurring low-accuracy points, such as the one in the northwest corner, possibly the result of cell-phone tower triangulation.",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! This issue is also visualised in Figure X. The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing data and measurement error in secondary location logs. We will compare PPMI to similar solutions in the literature by evaluating the distance between points which were simualted as missing and their imputed  counterparts. Finally we will calculate time spent at home as a function of the imputation method.

@@

How have researchers dealt with missing data in human mobility logs thus far? Unfortunately there is no golden standard in how to deal with this type of missing data. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. This vagueness is worrisome as it invites solutions which contain significant researcher degrees of freedom [@simmons]. The vaugeness is possibly also due to the fact that most researchers are unfamiliar with possible solutions. Most researchers simply downsample temporally and remove missing observations or use some sort of rule-based common sense imputations (e.g. @palmius_detecting_2017). The only principled approach that we know of that aims to solve the issue of missing data in location logs as they relate to human mobility is that of @barnett_inferring_2016. We will explore the methods of @barnett_inferring_2016  and @palmius_detecting_2017 in detail subsequently, after introducing exploring other spatiotemporal methods.

@@

For secondary logs the main limitation of SSM implementations is that they ignore movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. In other words, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. Moreover, these models often assume time and space invariance (location is not a direct function of time or space). These mathematical assumptions are violated in the case of human movement patterns. For this reason we do not consider existing SSMs to be useful for imputing missing data in this case.

In the wider realm of spatiotemporal statistics there are numerous missing data imputation methods. These often come from climate or geological research and rely on spatiotemporal autocorrelations. For instance, the CUTOFF method estimates missing values by incorporating similar observed temporal information from the value’s nearest spatial neighbors [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations). For this reason they cannot be easily applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns [ @palmius_detecting_2017 ;@barnett_inferring_2016; @Wu]. Importantly, none of them worked with secondary logs. Nonetheless we will detail what they did as informative examples. @palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

Missing data was imputed using the mean of measurements close in time if the participant was recorded within 500m of either end of a missing section and the missing section had a length of $\leq 2h$ or $\leq 12h$ after 9pm. In cases where the previous conditions are not met no values are imputed.

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by first converting data to mobility traces, which are defined as a sequence of flights and pauses. Flights are segments of linear movements and pauses corresponding to periods of time where a person does not move. Subsequently, the authors impute missing data by:

@@

On the other hand, @wu use what they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict future human movement. While the authors are concerned with prediction and not imputation, they devised a method called the Spatial Temporal Semantic (STS) algorithm which converts raw measurements to machine learning friendly discrete bins. Working with high-frequency measurements, @wu's method downsamples the raw data temporally and map-matches the resulting bins to discrete points along pre-established geographical features such as roads and highways. This minimises measurement error and paves the way for applying machine learning methods to human mobility problems.

In this section we have argued that there is a lack of established practices to follow with respect to missing data in human mobility logs. Moreover, extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns in secondary logs. Finally we discussed in detail three approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016; @wu].

@@

Subsequently let us simplify by assuming that a persons location is on two-dimensional Euclidian plane. Let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. For simplicity, we can assume that the x-axis is the longitude and the y-axis is the lattitude. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing the lattitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data. Furthermore, $G(t)$, $D$ and $a$ are indexed by time labled by the countable set $t = t_1 < ... < t_{n+1}$. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

@@

### Personalised Map Matching Imputation

Our algorithm is designed to leverage the periodic nature of human movement along with the long span of secondary to deal with measurement sparcity and inaccuracy.

@@

### Personalised Map Matching Imputation algorithm

@@

1. Map building: Extract from measurements $D$ all pause locations and path clusters to create a personalised map.
2. Binning: Assign each measurement $D$ to a unique discrete location. 
3. Imputing: Use a classification method to predict missing measurements based on the time as well as the distance in time to the next and previous measurements.

#### Map building

Following @wu's spatial-temporal-semantic (STS) feature extraction algorithm our aim is to transform pause and path locations into machine learning friendly discrete location sequences. There are multiple ways of extracting such measurement clusters in the literature, such as Spatio-Temporal Density-Based Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence oriented clustering (SOC) [@xiang;]. We will focus on two methods which explicitly with mobility patterns from unevenly sampled smartphone logs [ @palmius_detecting_2017 ;@barnett_inferring_2016]. Both of these methods pre-process the data and subsequently use two steps to extract pause locations: first they extract pauses and their corresponding locations, then they cluster pause locations based on spatial proximity. For simplicity this implementation of PMMI uses a stricter version of @barnett_inferring_2016's approach to extract pauses. However, there is no reason why more sophisticated methods cannot be used.

@@

1. The next measurement $D_{t+1}$ is within $t_{\text{Pause lim}}$ amount of seconds (so it is not missing)
2. The next measurement $D_{t+1}$ is within $d_{\text{Pause lim}}$ meters.
3. The duration of the pause is more than $\delta_{\text{Pause lim}}$ seconds.
4. Let the measurements of a possible pause which fit the aforementioned criteria be $D_{t,t+1,...,t+n}$. These points are only a pause if the distance between the mean coordinates of $D_{t,t+1,...,t+n}$ and the furthest away points of $D_{t,t+1,...,t+n}$ is within 2 times the mean accuracy $a$ of $D_{t,t+1,...,t+n}$.

@@

For all remaining points we assume that they belong to paths. In this implementation of PMMI we use the following algorithm to estimate paths:

1. Take all measurements which are not pauses, filter them based on an accuracy threshold $a_{\text{Path Lim}}$.
2. Create a distance matrix for all remaining measurements $D_t \in F$ and hierachically cluster it accordingly, such that all points within $d_{\text{Path Lim}}$ meters of each other are clustered into a single pause point.

At this point all empirically observed path clusters and pause clusters are extracted. However, there may be some overlap between pause clusters and path clusters. Thus, the clusters are clustered again, such that the pause clusters retain priority. This means that if a pause cluster and a path cluster are within less than $d$ meters of each other, the path cluster was removed. The reasoning for this is that the threshold for not being in a pause cluster should be higher, as individuals spend the majority of time at a pause cluster. The end result is a discretised map which contains pause and flight clusters based on the history of the individual.

### Binning

@wu's spatial-temporal-semantic (STS) feature extraction algorithm uses map matching as a ground truth to assign noisy measurements into discrete bins along roads. In other words, in addition to the measured data they also use a geographic database that contains information about the area in which the individual is (e.g. where precisely the roads are), and sort measurements into bins based on both the measurement and the geographic data base. For example, if an individual is measured as moving closely in parallel to a road A in an area where there is no other parallel road, @wu's method will assume that the individual is on the road A.

The key modification in PMMI is that whilst @wu uses data from outside the persons location logs, we use the total location history of the individual to create a personalised map which will be then used to bin measurements. This is feasible for two reasons: humans tend to have repetitive movement habits and secondary logs tend to be long. To put it in simpler terms, we consider each measurement at $D_x$ as a sample of $Path_{ab}$, and by aggregating many measurements we can use them to map out $Path_ab$.

Thus, all measurements $D$ were are then assigned to a discrete bin on the personal map. This includes previous measurements which were discarded from the map building excercise due to an accuracy $a$ value which exceeded $a_{\text{P lim}}$ or $a_{\text{F lim}}$. Again, for simplicity we used a simple assigning function, whereby the measurements where assigned to the bin nearest to the measurement.

#### Classification

At this point the objective of PMMI is to take all the information available about the mobility history of an individual and impute the missing value. In this implementation we trained an artificial neural network (ANN) to do so. The input variables to the ANN are:

1. The previous and subsequent observed bin.
2. The distance in time to the next & previous bin.
3. The time of the day encoded as a cyclical two-dimensional feature.
4. The day of the month as a binary class matrix.
5. The month of the year as a binary class matrix.

For the encoding of the time of day we took the cosine and the sine transforms of the amount of seconds that have elapsed after midnight [@london]. This is necessary so that the model can understand that one second past midnight and one second before midnight are in fact two seconds away from each other.

@@

For a missing timepoint at $D_t \in \emptyset$, the output of the model is a set of probability estimates associated with every location cluster. That is, for each missing timepoint the model returns a vector of probability estimates (with one estimate per cluster) associated with where the individual is.
